United States Public Elementary-Secondary Education spending (2016)
========================================================
by Z. McLaughlin
========================================================

# Introduction

There's been a lot of media attention to teacher salaries.  This exploration
covers various elements of how money is spent on education.  One area of 
particular interest is spending in Oklahoma which has gotten a lot of attention
in the news recently.  Exploration covers overall trends in spending per student
across the country, but then focuses is in Oklahoma vs California vs New Jersey
to see how things compare.

[Example youtube video of teachers in Oklahoma changing jobs.](http://kfor.com/2018/06/06/watch-oklahoma-teachers-break-down-explaining-why-they-are-leaving-the-classroom/)

# Dataset

There is a government census done of all public schools in the United States.
This data for 2016 is published here:

[2016 Public Elementary-Secondary Education Finance Data](https://www.census.gov/data/tables/2016/econ/school-finances/secondary-education-finance.html)

Since none of the provided tables contained all the desired information created
a cleaned dataset using the following information:

* [All Data Items](https://www2.census.gov/programs-surveys/school-finances/tables/2016/secondary-education-finance/elsec16.xls) - Contains the most comprehensive data.
* [elsec16t.xls](https://www2.census.gov/programs-surveys/school-finances/tables/2016/secondary-education-finance/elsec16t.xls) - Contains state information for each school
* [school16doc.doc](https://www2.census.gov/programs-surveys/school-finances/tables/2016/secondary-education-finance/school16doc.doc) - Contains the key for all the column names and key for states. 

Tables attached to project assignment:

* elsec_16.csv
* elect16.csv
* state_info_table.csv

Cleaned data created has the following column headers:

* "STATE" - Two letter state abbreviation
* "IDCENSUS" - Census id for the school district
* "NAME" - School district name
* "CONUM" - ANSI State and County Code
* "ENROLL" - Number of students enrolled
* "TOTALREV" - Total revenue in thousands
* "TFEDREV" - Total federal revenue in thousands
* "TSTREV" - Total state revenue in thousands
* "TLOCREV" - Total local revenue in thousands
* "TOTALEXP" - Total Expenses in thousands
* "TCURISAL" - Total salaries & wages for instruction in thousands
* "TCURIBEN" - Total benefits for instruction in thousands
* "PPCSTOT" - Per student total spending
* "PPSALWG" - Per student total salaries and wages
* "PPEMPBEN" - Per student total benefit payments
* "PPITOTAL" - Per student total spending for instruction
* "PPISALWG" - Per student total spending for salaries and wages for instruction
* "PPIEMBEN" - Per student total employee benefits for instruction

After cleaning data, decided to only include school districts with at least
50 students enrolled.

Added a few more columns later in the exploration:

* "PPTISB" - Per student money spent on salaries and benefits for instruction
* "PPPOI" - Percentage of money spent on teacher compensation
* "PLOCALREV" - Local rev percentage of total revenue
* "OKCA" - Limited number of states for analysis
* "PPNORM" - Teacher compensation per student normalized for cost 
of living.
* "PPTEACH" - Teacher compensation per student normalized for class size

Note:  Teacher compensation = Salary + benefits

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(dplyr)
library(data.table)
library(GGally)
```

```{r echo=FALSE,message=FALSE, warning=FALSE, results="hide",load_the_data}

# This section loads and cleans up the data


# This is the big file that contains lots of key info, but for some reason does not 
# contain state info so will need to merge with a file that contains state info. 

full_sd <- read.csv('./data/elsec_16t.csv')
head(full_sd)

# Cleaning up the data to just the data of interest

sdw <- subset(full_sd, 
              select=c(
                "IDCENSUS", 
                "NAME",
                "CONUM",
                "ENROLL",
                "TOTALREV",
                "TFEDREV",
                "TSTREV",
                "TLOCREV",
                "TOTALEXP",
                "TCURISAL",
                "TCURIBEN",
                "PPCSTOT",
                "PPSALWG",
                "PPEMPBEN",
                "PPITOTAL",
                "PPISALWG",
                "PPIEMBEN"
                ))
head(sdw)

# Discovered some data inconsistencies when cleaning the data and found looking 
# for Calhoun reveals some problems when the data was merged incorrectly so used
# searches for Calhoun to spot check the data is correct. Commented these lines
# out laster because knitr seemed to have a problem with them.  

# sdw[sdw$NAME %like% "CALHOUN", ]

# Reading in table that contains state info.

stateid <- read.csv('./data/elsec16.csv')
head(stateid)
stateid <- subset(stateid, 
              select=c(
                "STATE",
                "IDCENSUS",
                "NAME"
              ))
head(stateid)

stateid[stateid$NAME %like% "CALHOUN", ]

# Reading in the state data tables to take state from numerica to two letter 
# abreviation.

state_name <- read.csv('./data/state_info_table.csv')

head(state_name)

# Discovered the state code was in two different formats.  Using help from
# https://stackoverflow.com/questions/40097343/joining-two-datasets-with-different-classes
# was able to fix them up to match them up..
sapply(state_name,class)
sapply(stateid,class)

state_name$State.Code <- as.integer(as.character(state_name$State.Code))

# Joining tables to add in the state abbreviation.

newone <- stateid %>%
  left_join(state_name, by= c("STATE" = "State.Code"))
head(newone)

# newone[newone$NAME %like% "CALHOUN", ]

# Verifying that IDCENSUS have same type in different tables.
sapply(newone,class)
sapply(sdw,class)

full_working <- newone %>%
  left_join(sdw, by= c("IDCENSUS" = "IDCENSUS"))

head(full_working)

# full_working[full_working$NAME.x %like% "CALHOUN", ]

# Need to reduce down to just desired data.  There's a lot of data not needed
# For this investigation.

colnames(full_working)

full_working_subset <- subset(full_working, 
                       select=c(
                         "State",
                         "IDCENSUS",
                         "NAME.y",
                         "CONUM",
                         "ENROLL",
                         "TOTALREV",
                         "TFEDREV",
                         "TSTREV",
                         "TLOCREV",
                         "TOTALEXP",
                         "TCURISAL",
                         "TCURIBEN",
                         "PPCSTOT",
                         "PPSALWG",
                         "PPEMPBEN",
                         "PPITOTAL",
                         "PPISALWG",
                         "PPIEMBEN"
                       ))

head(full_working_subset)
# full_working_subset[full_working_subset$NAME.y %like% "CALHOUN", ]

# Need to change some of the column names make it all consistent.

colnames(full_working_subset)[colnames(full_working_subset)=="NAME.y"] <- "NAME"
colnames(full_working_subset)[colnames(full_working_subset)=="State"] <- "STATE"

colnames(full_working)

# Checking to make sure the merged data looks good

#full_working_subset[full_working_subset$NAME %like% "CALHOUN", ]

# Preliminary investigation shows there are schools with 0 enrollment.  Decide 
# only include schools with 50+ students Enrolled.

summary(full_working_subset)
clean_data <- full_working_subset[full_working_subset$ENROLL>=50,]
clean_data <- clean_data[clean_data$ENROLL != "NA",]


head(clean_data)
# clean_data[clean_data$NAME %like% "CALHOUN", ]

summary(clean_data)
write.table(clean_data, "./data/clean_edu_data.csv", sep="\t")



```


```{r echo=FALSE, message=FALSE, warning=FALSE, setting_variable_info}

# Setting up variables to use in graphs for expanded info on what is being
# graphed.

ENROLL_SUB <- "Number of students enrolled in the district"
TOTALREV_SUB <- "Total revenue in thousands (USD)"
TFEDREV_SUB <-"Total federal revenue in thousands (USD)"
TSTREV_SUB <- "Total state revenue in thousands (USD)"
TLOCREV_SUB <-  "Total local revenue in thousands (USD)"
TOTALEXP_SUB <- "Total Expenses in thousands (USD)"
TCURISAL_SUB <-"Total salaries & wages for instruction in 
  thousands(USD)"
TCURIBEN_SUB <- "Total benefits for instruction in thousands (USD)"
PPCSTOT_SUB <- "Per student total spending (USD)"
PPSALWG_SUB <- "Per student total salaries and wages (USD)"
PPEMPBEN_SUB <- "Per student total benefit payments (USD)"
PPITOTAL_SUB <- "Per student total spending for instruction (USD)"
PPISALWG_SUB <- "Per student total spending for salaries and wages for   
instruction (USD)"
PPIEMBEN_SUB <- "Per student total employee benefits for instruction (USD)"

# The columns that are added later

PPTISB_SUB <- "Teacher compensation per student (USD)"
PPPOI_SUB <- "Percentage of money spent on intr sal and benefits (0-1)"
PLOCALREV_SUB <- "Local rev percentage of total revenue (0-1)"
OKCA_SUB <- "New column with limited number of states for analysis"
PPNORM_SUB <- "Teacher comp per student normalized for cost of living (USD)"
PPTEACH_SUB <- "Teacher comp per student normalized for class size (USD)"
```

#  Summary of cleaned data

```{r echo=FALSE,message=FALSE,warning=FALSE,clean_data_summary}

sdata <- subset(clean_data, 
                       select=c(
                         "ENROLL",
                         "TOTALREV",
                         "TFEDREV",
                         "TSTREV",
                         "TLOCREV",
                         "TOTALEXP",
                         "TCURISAL",
                         "TCURIBEN",
                         "PPCSTOT",
                         "PPSALWG",
                         "PPEMPBEN",
                         "PPITOTAL",
                         "PPISALWG",
                         "PPIEMBEN"
                       ))
summary(sdata)

```
The data varies a lot:  

* Enrollment - One school district has almost a million students while others 
have less than 50 (although anything less than 50 was filtered out.)
* Total revenue max is over $27 billion dollars 
* Max spent per pupil/student is $228,102
* Just a quick look at the Revenue from federal, state and local, it's clear
that federal only plays a small part in funding education
* The per student total spending varies from -1,472 to 228,102
indicating that 1.  Some data might be questionable so it's better to look
at trends rather than the outliers.

# Univariate Plots Section
> **Histogram exploration**

Investigating how large the school districts are (note each district may 
contain a different number of separate schools.)  Also investigating what 
the per student spending looks like. Median is marked with red dashed line.
```{r echo=FALSE,message=FALSE,warning=FALSE,histograms_of_spending}

# Setting text size samller to make the graphs more readable
theme_set(theme_grey(base_size = 10)) 


ggplot(data=clean_data, aes(x=ENROLL )) +
  xlim(0,80000) + 
  xlab(ENROLL_SUB) + 
  geom_histogram(binwidth=500,color='black',fill='blue') + 
  geom_vline(xintercept=median(clean_data$ENROLL), linetype="dashed",
             color='red')

# Because the above graph is right skewed trying log 10.  

ggplot(data=clean_data, aes(x=ENROLL)) +
  ggtitle("Histogram using log 10 - note 'x' axis scale") +
  geom_histogram(color='black',fill='grey') +
  scale_x_log10()

# Box plot to get a better idea of the distribution.
qplot(x="",y=ENROLL,
  data =clean_data,
  geom = 'boxplot') +
  xlab("") + 
  coord_cartesian(ylim = c(0,10000))



ggplot(data=clean_data, aes(x=TOTALREV )) +
  xlim(0,500000) + 
  xlab(TOTALREV_SUB) + 
  geom_histogram(binwidth=10000,color='black',fill='yellow') +
  geom_vline(xintercept=median(clean_data$TOTALREV), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=TOTALEXP) ) +
  xlim(0,500000) + 
  xlab(TOTALEXP_SUB) + 
  geom_histogram(binwidth=10000,color='black',fill='orange') + 
  geom_vline(xintercept=median(clean_data$TOTALEXP), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=TFEDREV) ) +
  xlim(0,80000) + 
  xlab(TFEDREV_SUB) + 
  geom_histogram(binwidth=1000,color='black',fill='white') + 
  geom_vline(xintercept=median(clean_data$TFEDREV), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=TSTREV) ) +
  xlim(0,80000) + 
  xlab(TSTREV_SUB) + 
  geom_histogram(binwidth=1000,color='black',fill='pink') + 
  geom_vline(xintercept=median(clean_data$TSTREV), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=TLOCREV) ) +
  xlim(0,80000) + 
  xlab(TLOCREV_SUB) +
  geom_histogram(binwidth=1000,color='black',fill='light blue') +
  geom_vline(xintercept=median(clean_data$TLOCREV), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=PPCSTOT) ) +
  xlim(0,30000) + 
  xlab(PPCSTOT_SUB) + 
  geom_histogram(binwidth=500,color='black',fill='green') + 
  geom_vline(xintercept=median(clean_data$PPCSTOT), linetype="dashed",
             color='red') 


# Box plot to get a better idea of the distribution.
qplot(x="",y=PPCSTOT,
  data =clean_data,
  geom = 'boxplot') +
  xlab("") + 
  coord_cartesian(ylim = c(5000,25000))

ggplot(data=clean_data, aes(x=PPITOTAL) ) +
  xlim(0,30000) + 
  xlab(PPITOTAL_SUB) + 
  geom_histogram(binwidth=500,color='black',fill='brown') +
  geom_vline(xintercept=median(clean_data$PPITOTAL), linetype="dashed",
             color='red')

ggplot(data=clean_data, aes(x=PPISALWG) ) +
  xlim(0,15000) + 
  xlab(PPISALWG_SUB) +
  geom_histogram(binwidth=500,color='black',fill='magenta') + 
geom_vline(xintercept=median(clean_data$PPISALWG), linetype="dashed",
           color='red')

ggplot(data=clean_data, aes(x=PPIEMBEN) ) +
  xlim(0,15000) + 
  xlab(PPIEMBEN_SUB) + 
  geom_histogram(binwidth=500,color='black',fill='purple') +
  geom_vline(xintercept=median(clean_data$PPIEMBEN), linetype="dashed",
             color='red')


```


* All the histograms are right skewed with there still being values far to the 
right.
* Using a histogram with log 10 was able to see the normalized distribution for 
enrollment.  This was a good sest of data for using the log 10 because it's
easy to think of numbers of people in sets of 100,1000,10000, etc.
* Using a boxplot it was possible to see that although there's a large variation 
in school size, there's a large number of school districts under 2500 students.
* The total amount spent per student is also skewed right, but a little more 
normalized.  It is unexpected that the amount spent on a per student basis would
differ so drastically. Possible factors: cost of living, special needs 
population?
* Also created a box plot for total spending per student to get an idea of
the distribution.

# Bivariate Plots 

> **GGPAIRS matrix** - Getting some basic correlation data

```{r echo=FALSE,message=FALSE,warning=FALSE,ggpairs}

# Trimming to get easy to read matrix plot
mdata <- subset(clean_data, 
                       select=c(
                         "ENROLL",
                         "TOTALEXP",
                         "PPCSTOT",
                         "PPITOTAL",
                         "PPISALWG",
                         "PPIEMBEN"
                       ))


ggpairs(mdata)

```

> **Plots of enrollment vs expenses**

```{r echo=FALSE,message=FALSE,warning=FALSE,enroll_vs_expense}

ggplot(data=clean_data, aes(x=ENROLL,y=TOTALEXP)) +
  xlim(0,125000) + 
  ylim(0,2000000) +
  xlab(ENROLL_SUB) + 
  ylab(TOTALEXP_SUB) + 
  ggtitle("Includes schools up to 120,000 students") + 
  geom_point(color='blue',alpha=1/4)

ggplot(data=clean_data, aes(x=ENROLL,y=TOTALEXP)) +
  xlim(0,10000) + 
  ylim(0,750000) +
  xlab(ENROLL_SUB) + 
  ylab(TOTALEXP_SUB) + 
  ggtitle("Includes schools up to 10,000 students") +
  geom_point(color='dark green',alpha=1/4)

```

The relationship between enrollment and and total expenses looks very linear at 
the lower numbers, but as the enrollment goes up, the relationship
scatters.  Also even though the relationship is fairly linear there is quite
a bit of variation and some outliers.

Note that at this point decided to look mainly at  districts with less than 
125,000 vs districts like NYC that has almost a million students or Hawaii where
the whole state is one district.

> **Plots of enrollment vs total per student sending**

```{r echo=FALSE,message=FALSE,warning=FALSE,enroll_vs_ppexpense}

ggplot(data=clean_data, aes(x=ENROLL,y=PPCSTOT)) +
  xlim(0,125000) + 
  ylim(0,50000) +
  geom_point(color="blue",alpha=1/5) + 
  xlab(ENROLL_SUB) + 
  ylab(PPCSTOT_SUB) + 
  ggtitle("Includes schools up to 125,000 students") +
  geom_hline(yintercept=median(clean_data$PPCSTOT), linetype="dashed",color='red')


ggplot(data=clean_data, aes(x=ENROLL,y=PPCSTOT)) +
  xlim(0,10000) + 
  ylim(0,50000) +
  geom_point(color="dark green",alpha=1/5) + 
  xlab(ENROLL_SUB) + 
  ylab(PPCSTOT_SUB) + 
  ggtitle("Includes schools up to 50,000 students") +
  geom_hline(yintercept=median(clean_data$PPCSTOT), linetype="dashed",color='red')


ggplot(data=clean_data, aes(x=ENROLL,y=PPCSTOT)) +
  xlim(0,1000) + 
  ylim(0,50000) +
  xlab(ENROLL_SUB) + 
  ylab(PPCSTOT_SUB) + 
  ggtitle("Includes schools up to 1,000 students") +
  geom_point(color="orange",alpha=1/5) + 
  geom_hline(yintercept=median(clean_data$PPCSTOT), linetype="dashed",color='red')


```

Horizontal line on plots above is the median.  It appears that the spending per 
student is at the lower enrollments hovers around the median, but there are
quite a few schools where the spending is substantially higher. Starting to 
bring up questions:

* What is the difference between the school that spends $7,000/student vs the one
that spends $20,000/student?
* The giant school districts seem to hover closer to the median than the smaller
school districts. Why?


> **Plots of enrollment vs revenue sources (Federal, State, Local)**

```{r echo=FALSE,message=FALSE,warning=FALSE,enroll_vs_tax_revenue}

# Graphs of revenue sources.

ggplot(data=clean_data, aes(x=ENROLL,y=TFEDREV)) +
  xlim(0,125000) + 
  ylim(0,50000) +
  xlab(ENROLL_SUB) + 
  ylab(TFEDREV_SUB) + 
  geom_vline(xintercept=10000, linetype="dashed",color='red') + 
  geom_point(color="blue",alpha=1/5) 

ggplot(data=clean_data, aes(x=ENROLL,y=TSTREV)) +
  xlim(0,125000) + 
  ylim(0,1000000) +
  xlab(ENROLL_SUB) + 
  ylab(TSTREV_SUB) +
  geom_vline(xintercept=10000, linetype="dashed",color='red') +
  geom_point(color="dark green",alpha=1/5) 


ggplot(data=clean_data, aes(x=ENROLL,y=TLOCREV)) +
  xlim(0,125000) + 
  ylim(0,200000) +
  xlab(ENROLL_SUB) + 
  ylab(TLOCREV_SUB) +
  geom_vline(xintercept=10000, linetype="dashed",color='red') +
  geom_point(color="orange",alpha=1/5) 


```

Added a red dashed line for enrollment = 10,000 students to make it easier to compare
the various sources of funding.  

Looks like most money comes from
Where the money comes from looks interesting.  While state funding looks pretty 
linear.  Looks like most money for education comes from state and local sources.

# Bivariate Plots Section 2

Based on the plots above there is a need to create additional data to get more 
specific plots.

* PPTISB = PPISALWG + PPIEMBEN - Teacher compensation per student 
for instruction
* PPPOI = PPTISB/PPCSTOT - Percentage of money spent on teacher compensation 
* PLOCALREV = TLOCREV/TOTALREV - Local rev percentage of total revenue
* OKCA = OK, CA, NY, NJ, AZ, WA, OR, MA, IA OTHER to be able to more detailed 
information at the state level on the same plot.


```{r echo=FALSE,message=FALSE,warning=FALSE,new_columns_1}

# Adding additional computed variables for additional graphing.
clean_data$PPTISB = clean_data$PPISALWG + clean_data$PPIEMBEN
clean_data$PPPOI = clean_data$PPTISB/clean_data$PPCSTOT
clean_data$PLOCALREV= clean_data$TLOCREV/clean_data$TOTALREV
# head(clean_data)

summary(subset(clean_data, 
                       select=c(
                         "PPTISB",
                         "PPPOI",
                         "PLOCALREV"
                       )))


  

```

The summary showed percentages over 1 for percent of total spent per pupil for 
salaries, wages benefits.  Did some spot checks and found that there were only a
handful of schools impacted.  Brings up questions as to the reliability of the 
data, but since this is not for professional use, will not investigate further.



```{r echo=FALSE,message=FALSE,warning=FALSE,new_columns_2}

# Creating new column with 9 states plus other 

clean_data$OKCA = "Other"
clean_data[clean_data$STATE == "CA",]$OKCA <- "CA"
clean_data[clean_data$STATE == "OK",]$OKCA <- "OK"
clean_data[clean_data$STATE == "NY",]$OKCA <- "NY"
clean_data[clean_data$STATE == "AZ",]$OKCA <- "AZ"
clean_data[clean_data$STATE == "NJ",]$OKCA <- "NJ"
clean_data[clean_data$STATE == "OR",]$OKCA <- "OR"
clean_data[clean_data$STATE == "WA",]$OKCA <- "WA"
clean_data[clean_data$STATE == "IA",]$OKCA <- "IA"
clean_data[clean_data$STATE == "MA",]$OKCA <- "MA"

small_set <- clean_data[clean_data$OKCA != "Other", ]
#small_set[small_set$ENROLL == "NA",]

just_three <- subset(small_set, OKCA == "CA" | OKCA == "NJ" | OKCA == "OK")

```
> **Plots Looking at percentage of local rev vs other variables**

```{r echo=FALSE,message=FALSE,warning=FALSE,per_local_rev}

ggplot(data=clean_data, aes(x=TOTALEXP,y=PLOCALREV)) +
  xlim(0,250000) +
  xlab(TOTALEXP_SUB) + 
  ylab(PLOCALREV_SUB) + 
  ggtitle("Includes revenue up to 250,000,000 dollars") +
  geom_point(color="black",alpha=1/5) 

ggplot(data=clean_data, aes(x=PPCSTOT,y=PLOCALREV,alpha=1/100)) +
  xlim(0,50000) +
  xlab(PPCSTOT_SUB) + 
  ylab(PLOCALREV_SUB) + 
  ggtitle("Includes up to 50,000 dollars per student spending") +
  geom_point(color="blue",alpha=1/5) 


ggplot(data=clean_data, aes(x=PPTISB,y=PLOCALREV)) +
  xlim(0,20000) +
  xlab(PPTISB_SUB) + 
  ylab(PLOCALREV_SUB) + 
  ggtitle("Includes instructional salary and benefits per student of 20,000") +
  geom_point(color="red",alpha=1/5) 


```

Had expected to maybe be able to see a trend, but did not see anything useful.
Decided to look at a subset of data with a limited number of states.

# Multivariate Plots Section

> **Plots Looking at data for a subset of states**


```{r echo=FALSE,message=FALSE,warning=FALSE,subset_of_states_1}

ggplot(data=small_set, aes(x=ENROLL,y=TOTALEXP)) +
    facet_wrap(~ OKCA) +
    xlim(1000,10000) +
    ylim(0,100000) +
    xlab(ENROLL_SUB) + 
    ylab(TOTALEXP_SUB) +
    ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point(color="black",alpha=1/2) 

ggplot(data=small_set, aes(x=ENROLL,y=PLOCALREV)) +
   facet_wrap(~ OKCA) +
    xlim(1000,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PLOCALREV_SUB) +
    ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point(color="blue",alpha=1/2) 

ggplot(data=small_set, aes(x=ENROLL,y=PPTISB)) +
   facet_wrap(~ OKCA) +
    xlim(50,1000) +
    ylim(0,25000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
    ggtitle("Includes schools of 50 - 1000 students") +
  geom_point(color="orange",alpha=1/2) 

ggplot(small_set, aes(x=ENROLL,y=PPTISB)) +
   facet_wrap(~ OKCA) +
    xlim(1000,10000) +
    ylim(0,25000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
      ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point(color="red",alpha=1/2) 


ggplot(data=small_set, aes(x=ENROLL,y=PPPOI)) +
   facet_wrap(~ OKCA) +
    xlim(1000,10000) +
    ylim(.25,.75) +
    xlab(ENROLL_SUB) + 
    ylab(PPPOI_SUB) +
    ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point(color="green",alpha=1/2) 
```

Starting to see some noticeable differences:

* New York spends a lot more on education than Oklahoma.  
* California has a lot of variation in spending.
* It's looking pretty clear that MA, NY, and NJ are spending significantly more 
on teacher compensation than CA, AZ, OK.

> **Plot Looking at just 1 state (New York) a little deeper **

```{r echo=FALSE,message=FALSE,warning=FALSE,subset_of_states_2}
ggplot(data=small_set[small_set$OKCA=="NY",], 
       aes(x=ENROLL,y=PPCSTOT,color=PLOCALREV)) +
    xlim(1000,10000) +
    ylim(0,30000) +
    xlab(ENROLL_SUB) + 
    ylab(PPCSTOT_SUB) +
  geom_point() +
  scale_colour_gradientn(colours=rainbow(4)) 

```

It does appear that total per student spending goes up when the percent of 
funding from local revenue for the school district goes up.

> **Narrowing down to instructional salary and benefits **

* "PPISALWG" - Per student total spending for salaries and wages for instruction
* "PPIEMBEN" - Per student total employee benefits for instruction

Looking at a few more factors.

```{r echo=FALSE,message=FALSE,warning=FALSE,more_plots_dollars_to_sb}
ggplot(data=small_set, aes(x=ENROLL,y=PPISALWG)) +
   facet_wrap(~ OKCA) +
    xlim(50,10000) +
    ylim(2500,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPISALWG_SUB) +
  geom_point(color="red") 

ggplot(data=small_set, aes(x=ENROLL,y=PPIEMBEN)) +
   facet_wrap(~ OKCA) +
    xlim(50,10000) +
    ylim(0,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPIEMBEN_SUB) +
  geom_point(color="orange") 

ggplot(small_set, aes(x=ENROLL,y=PPTISB)) +
   facet_wrap(~ OKCA) +
    xlim(50,10000) +
    ylim(2500,20000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
  geom_point(color="green")

```


> **Narrowing down to just 3 states, NJ, CA, & OK **

```{r echo=FALSE,message=FALSE,warning=FALSE,more_plots_3_variables_}
ggplot(just_three, aes(x=ENROLL,y=PPTISB,color=OKCA)) +
    xlim(50,1000) +
    ylim(2500,11000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
    ggtitle("Includes schools of 50 - 1000 students") +
  geom_point()


ggplot(just_three, aes(x=ENROLL,y=PPTISB,color=OKCA)) +
    xlim(1000,10000) +
    ylim(2500,11000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
   ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point()

ggplot(just_three, aes(x=ENROLL,y=PPCSTOT,color=OKCA)) +
    xlim(50,1000) +
    ylim(2500,20000) +
    xlab(ENROLL_SUB) + 
    ylab(PPCSTOT_SUB) +
    ggtitle("Includes schools of 50 - 1000 students") +
  geom_point()


ggplot(just_three, aes(x=ENROLL,y=PPCSTOT,color=OKCA)) +
    xlim(1000,10000) +
    ylim(2500,20000) +
    xlab(ENROLL_SUB) + 
    ylab(PPCSTOT_SUB) +
    ggtitle("Includes schools of 1000 - 10,000 students") +
  geom_point()

# head(just_three)
```

CA is way below NJ and not too far above OK although it's pretty well known that 
it's much more expensive to live in California.  So what happens if a cost of 
living index is applied to the values?

> **Factoring in the cost of living for NJ, CA, & OK **

Grabbed some data from 2017 (close enough) from:
[US Learning](https://www.uslearning.net/cost-of-living-by-state.html)

* CA - 136.3
* NJ - 121.2
* OK - 89.1

To normalize the data will take the total spent on salaries and benefits
per pupil divided by the index/100 just to get a more dollar based amount.

```{r echo=FALSE,message=FALSE,warning=FALSE,normalized}

#just_three[just_three$STATE == "CA",]$OKCA <- "CA"
#head(just_three)

just_three$PPNORM[just_three$OKCA == "CA"] <-  just_three$PPTISB * (100/136.3)
just_three$PPNORM[just_three$OKCA == "NJ"] <-  just_three$PPTISB * (100/121.2)
just_three$PPNORM[just_three$OKCA == "OK"] <-  just_three$PPTISB * (100/89.1)

ggplot(just_three, aes(x=ENROLL,y=PPNORM,color=OKCA)) +
    xlim(50,1000) +
    ylim(1000,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPNORM_SUB) +
   ggtitle("Includes school districts of 50 - 1000 students") +
  geom_point()

ggplot(just_three, aes(x=ENROLL,y=PPNORM,color=OKCA)) +
    xlim(1000,10000) +
    ylim(1000,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPNORM_SUB) +
    ggtitle("Includes school districts of 1000 - 10,000 students") +
  geom_point()



```

Based on the graphs above accounting for cost of living, it seems Oklahoma 
teachers get paid more than either New jersey or California.

> **Factoring in the class size for NJ, CA, & OK **

In theory class size would have impact the actual salaries.  If compensation is 
the per student money spent on instructional salaries and benefits than:

* Higher class size = higher salary
* Lower class size = lower salary

If the per student instructional compensation were the same.

(Note: This doesn't examine quality of education based or difficulty of the 
teaching role as class size increases.)

From:  

[National Center for Education Statistics](https://nces.ed.gov/programs/digest/d16/tables/dt16_209.30.asp)

This information is from 2012 so things may be a little old.  Four years is a 
long time.  But unless the states increased or decreased disproportionately 
to other states the results should be similar.

Decide to use the numbers provided in the following way:

(Average class size by level of instruction Elementary + 
Average class size by level of instruction Secondary)  
divided by 2. Then normalize that using the average US class size

* CA - (25.0	+	32.0)/2  
* NJ - (18.5	+	23.9)/2	
* OK - (20.7	+	23.7)/2	
* US - (21.3	+	26.8)/2	 

```{r echo=FALSE,message=FALSE,warning=FALSE,class_size_normalized}
USS <- (21.3	+	26.8)/2	
CAS <- ((25.0	+	32.0)/2)/USS 
NJS <- ((18.5	+	23.9)/2)/USS	
OKS <- ((20.7	+	23.7)/2)/USS


just_three$PPTEACH[just_three$OKCA == "CA"] <-  just_three$PPTISB * CAS
just_three$PPTEACH[just_three$OKCA == "NJ"] <-  just_three$PPTISB * CAS
just_three$PPTEACH[just_three$OKCA == "OK"] <-  just_three$PPTISB * CAS

ggplot(just_three, aes(x=ENROLL,y=PPTEACH,color=OKCA)) +
    xlim(50,1000) +
    ylim(2500,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTEACH_SUB) +
    ggtitle("Includes school districts of 50 - 1000 students") +
  geom_point()

ggplot(just_three, aes(x=ENROLL,y=PPTEACH,color=OKCA)) +
    xlim(1000,10000) +
    ylim(2500,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTEACH_SUB) +
    ggtitle("Includes school districts of 1000 - 10,000 students") +
  geom_point()

```

**Results:**  Once you add in cost of living and class size there while there is 
a lot of variation across states, there doesn't appear to be a big difference
between California, New Jersey and Oklahoma in terms of Instructional Salaries
and benefits.

# Final Plots and Summary 

### Plot One 
```{r echo=FALSE, message=FALSE,warning=FALSE,Plot_One}
ggplot(just_three, aes(x=ENROLL,y=PPTISB,color=OKCA)) +
    xlim(1000,10000) +
    ylim(2500,20000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTISB_SUB) +
    labs(color="State") + 
    ggtitle("Data straight from published information",
            subtitle="Includes school districts of 1000 - 10,000 students") +
  geom_point()


```

### Description One
Using the cleaned up data from: 

[2016 Public Elementary-Secondary Education Finance Data](https://www.census.gov/data/tables/2016/econ/school-finances/secondary-education-finance.html)

to graph the per student spend on instructional salary and benefits for 
California, New Jersey and Oklahoma there is a clear indication that the per 
student spending in Oklahoma is far less than California or New Jersey for 
school districts between 1000 and 10,000 students.

It also appears that the variation of what is spent per student is far greater
in California and New Jersey than in Oklahoma.

Using this graph alone:

* Oklahoma teachers are justified in how upset they are regarding salary and
benefits provided.

### Plot Two
```{r echo=FALSE, message=FALSE,warning=FALSE,Plot_Two}
ggplot(just_three, aes(x=ENROLL,y=PPNORM,color=OKCA)) +
    xlim(50,1000) +
    ylim(1000,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPNORM_SUB) +
    labs(color="State") + 
  ggtitle("Data normalized for cost of living",
          subtitle="Includes school districts of 1000 - 10,000 students") +
  geom_point()  
```

### Description Two

Grabbed some data from 2017 (close enough) from:
[US Learning](https://www.uslearning.net/cost-of-living-by-state.html)

Normalized for cost of living:

* CA - 136.3
* NJ - 121.2
* OK - 89.1

The schools switch places.  California falls to the bottom.  New Jersey ends up 
in the middle and it appears that Oklahoma teachers are the best paid teachers.
Also the variation in spending per student on instructional salaries and 
benefits for Oklahoma appears larger once the cost of living index is applied 
since a dollar in Oklahoma goes farther than it does in California or 
New Jersey.


### Plot Three
```{r echo=FALSE, message=FALSE,warning=FALSE,Plot_Three}
ggplot(just_three, aes(x=ENROLL,y=PPTEACH,color=OKCA)) +
    xlim(1000,10000) +
    ylim(2500,10000) +
    xlab(ENROLL_SUB) + 
    ylab(PPTEACH_SUB) +
    labs(color="State") + 
    ggtitle("Data normalized for class size",
            subtitle="Includes school districts of 1000 - 10,000 students") +
  geom_point()

```

### Description Three

Because the data on plot 1 and plot 2 is NOT strictly reflective of how much 
a particular teacher gets paid in any of the states it is necessary to use a
normalizer to try to compute what the relationships might look like.

Using class size information:

From:  
[National Center for Education Statistics](https://nces.ed.gov/programs/digest/d16/tables/dt16_209.30.asp)

This information is from 2012 so things may be a little old.  Four years is a 
long time

Decide to use the numbers provided in the following way:

Average class size for teachers in self-contained classes	+ 
Average class size for teachers in departmentalized instruction	
and divide by 2. Then normalize that using the average US class size.

* CA - (25.0	+	32.0)/2  
* NJ - (18.5	+	23.9)/2	
* OK - (20.7	+	23.7)/2	
* US - (21.3	+	26.8)/2	 

and normalizing the data in plot 2 created the plot above.

Now the data from all three states starts to overlap indicating that maybe there
is less of a difference on what is spent on teacher salary and benefits by state
than what the original data showed alone.

Note that the final numbers on the y axis are not real numbers for any of the 
states, but an indication of spending relative to each other.

------

# Reflection

> **What were some of the struggles?**

  * I thought I had a really good dataset to start out with, but the more I 
  looked at it, the more I realized that I needed to clean it up first.
  * I had an expectation that different things about educational spending
  would pop out right away, but actually there is a lot of variation by school
  district, so it was harder to see trends than I expected.
  * I hadn't been using Rstudio previously in a way that could be used by knit
  so I had to learn a lot of the details of Rstudio while working on the 
  project.
  
> **What went well? **

  * Using the full dataset and a few different graphing techniques I was able to
  narrow down to an area of interest.
  * As I found some bad data in the dataset, I started to realize that I got 
  lucky that the dataset was as good as it was.
  
> **What was surprising?**

  * That government published data could have so much detailed information, but
  still have errors.
  * That the state of Hawaii has only one school district.
  * That NYC has almost a million students and a budget of 20+ billion
  dollars.
  * That there is so much variation on what is spent per student on across the
  country and across a state.
  * That there are so many individual school districts

> **What further investigations could be done?**

  * Could try to find more recent class size information and see if the results
  are the same.
  * Could could create data on a map.
  * Could do a more detailed analysis of just one state and how and why spending
  varies within a state.
  

